Predicting Conflict Escalation

### 2023-04-29 docs/hes_final/article.qmd
add umap figures and ffnn description

@@

Another direction is dimensionality reduction technique. We observed a clear separation of clusters using a
UMAP technique supervised by the Spike target variable. Using the dataset reduced with this technique
could be a better alternative to a principal component analysis our statistical model uses.

>>>>

## Feed Forward Neural Network

Our third approach employed a feed forward neural network model. Uniform manifold approximation and projection (UMAP) was used to reduce dimensionality of GDELT article embeddings in a supervised manner. Our spike variable was used to condition the dimension reduction process of the training data to encourage separation of the variable classes in the reduced embedding space. The fit UMAP model, was applied to the validation and test data without supervision. These reduced, 128 dimension embeddings, along with country context indicators, were used as inputs to a feed forward neural network with a single hidden layer of 32 neurons. The model's 4577 parameters were trained for 3 epochs.

![Distribution of supervised UMAP reduction to 2 dimensions](imgs/fig-umap2d-jointplot.png){width=50%}

![Scatterplot of 2d UMAP reduced articles with spike labels](imgs/fig-umap2d-ds-targets.png){width=50%}

Country context inputs consisted of two numerical indicators ('Under-five mortality rate 2019' and 'Adolescent population 2020 Proportion of total population (%) Total') and three one-hot encoded categorical indicators (3 categories of Human Development Index ranges, 4 categories of gross national income per capita, and 3 categories of Fragile States Index ranges).

```{mermaid}
%%| fig-width: 3
%%| fig-height: 5
%%| fig-cap: Fig 6. Feed forward neural network model architecture
flowchart TD
    GDELT[GDELT Embeddings] -->|sample by country by month| Articles(Monthly news article samples)

    Articles --> UMAP{Supervised UMAP}
    UMAP --> Compact(128 dimension representations)

    Compact --> FFNN{Feed Forward Neural Network}
    FFNN --> Pred(Prediction of conflict escalation in 3 months)

    ACLED[Monthly ACLED fatality counts] -->|compute our 'spike' variable| Spike
    Spike(Conflict response binary variable) --> |included during training|UMAP

    Spike -->FFNN

    Context[Country context indicators] --> FFNN
```

@@
